---
title: "Anatomy of the model: the generative process"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anatomy of the model: the generative process}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5, dpi = 96)
library(bayespulse)
```

An observed hormone series is the sum of a baseline level, pulsatile secretion
(a set of Gaussian pulses), exponential elimination back toward baseline, and
measurement error. This vignette builds intuition for that generative process
by simulating series and varying each piece in turn. The formal model, with its
secretion and elimination functions written out, is given in the getting-started
vignette; here the goal is simply to see what each parameter does to the shape
of the series.

## A single simulated series

We start from one realistic series to anchor the comparisons that follow.

```{r base-series}
set.seed(1)
sim <- simulate_pulse(num_obs = 72, interval = 10, mass_mean = 3.5,
                      width_mean = 35, constant_baseline = 2.6,
                      constant_halflife = 45)
plot(sim)
```

## Pulse mass

Pulse mass controls how much hormone is released per pulse: a larger mass
produces taller pulses.

```{r sweep-mass, fig.height = 7}
masses <- c(1.5, 3.5, 6.0)
op <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (m in masses) {
  set.seed(1)
  s <- simulate_pulse(num_obs = 72, interval = 10, mass_mean = m,
                      width_mean = 35, constant_baseline = 2.6,
                      constant_halflife = 45)
  plot(s$data$time, s$data$concentration, type = "l",
       main = paste("mass_mean =", m), xlab = "Time (min)", ylab = "Conc.")
}
par(op)
```

## Pulse width

Pulse width controls how spread out each secretion event is: a larger width
gives broader, flatter humps.

```{r sweep-width, fig.height = 7}
widths <- c(15, 35, 60)
op <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (w in widths) {
  set.seed(1)
  s <- simulate_pulse(num_obs = 72, interval = 10, mass_mean = 3.5,
                      width_mean = w, constant_baseline = 2.6,
                      constant_halflife = 45)
  plot(s$data$time, s$data$concentration, type = "l",
       main = paste("width_mean =", w), xlab = "Time (min)", ylab = "Conc.")
}
par(op)
```

## Half-life

The elimination half-life controls how quickly hormone clears between pulses:
a longer half-life means slower decay and higher troughs.

```{r sweep-halflife, fig.height = 7}
halflives <- c(20, 45, 90)
op <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (h in halflives) {
  set.seed(1)
  s <- simulate_pulse(num_obs = 72, interval = 10, mass_mean = 3.5,
                      width_mean = 35, constant_baseline = 2.6,
                      constant_halflife = h)
  plot(s$data$time, s$data$concentration, type = "l",
       main = paste("constant_halflife =", h), xlab = "Time (min)",
       ylab = "Conc.")
}
par(op)
```

## Inter-pulse interval

The mean inter-pulse interval controls how often pulses occur: a larger
interval yields fewer, more widely separated pulses.

```{r sweep-ipi, fig.height = 7}
ipis <- c(8, 14, 24)
op <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (ip in ipis) {
  set.seed(1)
  s <- simulate_pulse(num_obs = 72, interval = 10, ipi_mean = ip,
                      ipi_var = 10, ipi_min = 4, mass_mean = 3.5,
                      width_mean = 35, constant_baseline = 2.6,
                      constant_halflife = 45)
  plot(s$data$time, s$data$concentration, type = "l",
       main = paste("ipi_mean =", ip), xlab = "Time (min)", ylab = "Conc.")
}
par(op)
```

Together these four parameters, plus the baseline level and the measurement
error variance, generate the rich variety of shapes seen in real hormone
series. The next vignettes show how to encode prior knowledge about each of
them and how to fit the model to recover them from data.
