---
title: "Specifying priors and starting values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying priors and starting values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5, dpi = 96)
library(bayespulse)
```

In a Bayesian deconvolution model the priors encode what we know about the
hormone system *before* seeing the data. The observed series is decomposed into
a baseline, a set of pulses (each with a mass and a width), exponential
elimination, and measurement error. Many of these pieces are only weakly
identified from a single subject's data -- a tall narrow pulse and a shorter
wider one can produce similar-looking peaks, for example -- so reasonably
informative priors are not just a formality; they stabilize estimation and keep
the sampler in physiologically plausible regions.

`pulse_spec()` is where all of this is set: the priors themselves, the starting
values that seed the MCMC chains (the `sv_*` arguments), and the proposal
variances that tune Metropolis-Hastings steps (the `pv_*` arguments). This
vignette focuses on the priors and what they mean physiologically; starting
values and proposal-variance tuning are covered by the fitting and diagnostics
material. Because nothing here runs MCMC -- we only draw from priors and
simulate -- the examples are fast.

## What the priors mean

Each prior controls one part of the generative process. The means are on the
natural scale of the quantity; variances express how tightly that quantity is
constrained a priori.

| Argument | Controls | Typical choice / notes |
|---|---|---|
| `prior_baseline_mean`, `prior_baseline_var` | Constant baseline concentration | Center near the assay's observed trough; a moderate variance lets the data adjust it. |
| `prior_mass_mean`, `prior_mass_var` | Mean pulse mass (hormone released per pulse) | Set from prior studies of the hormone; mass is positive, so keep the mean comfortably above zero. |
| `prior_width_mean` | Mean pulse width (duration / spread of a pulse) | Reflects how long secretion events last; on the order of tens of minutes for many hormones. |
| `prior_sd_mass` | SD of pulse-to-pulse mass variability | Allows real pulses to differ in size; larger values admit more heterogeneity. |
| `prior_sd_width` | Pulse-to-pulse width variability | Fixed (not estimated) in this version; supplied for completeness. |
| `prior_halflife_mean`, `prior_halflife_var` | Elimination half-life | Often known fairly well from physiology, so a smaller variance is reasonable. |
| `prior_mean_pulse_count` | Expected number of pulses over the window | Anchors how many secretion events to expect; scale to the observation length. |
| `location_prior_type`, `prior_location_gamma`, `prior_location_range` | Spatial prior on pulse locations (e.g. `"strauss"`) | Repulsive prior; see "The location prior" below. |

## Informative vs vague priors

The variance arguments are the main dial between informative and vague. A small
prior variance encodes confidence -- the model is told a parameter is near a
particular value and the data have to work hard to move it (informative). A
large variance expresses ignorance and lets the data dominate (vague). The
right choice depends on how much trustworthy prior knowledge you have: half-life
is often well characterized and warrants a tighter prior, while baseline or
pulse mass may deserve more room.

A good way to see what a prior *implies* is a prior-predictive check: draw
parameters from the prior, simulate series, and look at the spread of outcomes.
Below we vary the baseline and half-life priors only, drawing each from a normal
and flooring at small positive values so `simulate_pulse()` always receives
valid inputs.

```{r prior-predictive, fig.height = 6}
set.seed(2026)
draw_series <- function(baseline_sd, halflife_sd) {
  bl <- max(0.3, rnorm(1, mean = 2.5, sd = baseline_sd))
  hl <- max(5,   rnorm(1, mean = 45,  sd = halflife_sd))
  simulate_pulse(num_obs = 60, interval = 10, mass_mean = 2.5,
                 width_mean = 35, constant_baseline = bl,
                 constant_halflife = hl)
}
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
# Informative priors: small SDs -> series cluster tightly
for (i in 1:2) {
  s <- draw_series(baseline_sd = 0.25, halflife_sd = 4)
  plot(s$data$time, s$data$concentration, type = "l",
       main = "informative prior", xlab = "Time (min)", ylab = "Conc.")
}
# Vague priors: large SDs -> series vary widely
for (i in 1:2) {
  s <- draw_series(baseline_sd = 1.2, halflife_sd = 25)
  plot(s$data$time, s$data$concentration, type = "l",
       main = "vague prior", xlab = "Time (min)", ylab = "Conc.")
}
par(op)
```

The two informative-prior draws (top row) look much alike: the prior tightly
constrains where the series can sit and how quickly it decays. The two
vague-prior draws (bottom row) range widely, because the prior offers little
constraint and very different baselines and decay rates are equally admissible.
Inspecting these prior-predictive series is a quick sanity check that your
priors imply plausible data before any fitting begins. This example varies only
the baseline and half-life for illustration, but the same logic applies to the
mass, width, and pulse-count priors.

## The location prior

Pulse locations get a spatial prior rather than independent draws. The default
`location_prior_type = "strauss"` uses a Strauss point process, which is
*repulsive*: it down-weights configurations in which pulses sit very close
together. This matches the biology -- real secretion events do not pile
arbitrarily close to one another, since the system needs time between bursts.
Two arguments tune it: `prior_location_gamma` is the repulsion strength (a value
of $\gamma = 1$ means no repulsion, so locations are effectively
order-independent, while smaller $\gamma$ imposes stronger repulsion that
discourages near-coincident pulses), and `prior_location_range` is the
interaction distance, the separation below which pulses are penalized.

## Where to go next

`population_spec()` mirrors these same prior ideas at the population level,
where the priors describe distributions of pulse parameters across subjects. The
getting-started vignette shows a complete `pulse_spec()` call in a fitted model,
and the diagnostics material covers starting values and convergence in detail.
